---
title: "AMF phylogeny and heatmap"
author: "Kian Kelly"
date: "Updated on 12/24"
output: html_document
---

#Fungal taxonomic barplot by site

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###STEP1: Load all necessary packages for analysis
More information about Phyloseq can be found at the following link: [Phyloseq](https://joey711.github.io/phyloseq/)
If you get error in this step, you probably need to install any packages which causes error.

```{r warning=FALSE, message=FALSE}
library(ape)
library(vegan)
library(dplyr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(plyr)
library(data.table)
library(tidyr)
library(tidyverse)
library(multcompView)
library(ggtree)
library(treeio)
library(devtools)
library(MicEco)
library(MoMAColors)
library(rphylopic)
library(cowplot)
```

###STEP2: Import Mapping file (metadate file)
1.Check mapping file before import to R, R doesn’t seem to like sample name to start with number or contain “-” in sample name. If you get error in this step, you should check file name first.

2.First column of first row should not start with #, R will not read the first row that starts with #

3. You can choose which samples to include in analysis by indicating "KEEP" in Description column

```{r}
meta = read.table("metadata.tsv",header=TRUE,row.names=1,sep="\t",stringsAsFactors=FALSE)
```

If need to keep or remove some samples
```{r}
#meta <- meta[which(meta$Layer %in% c("Surface")),]
```

###STEP3: Check if your metadata file has been import successfully and correctly

The output will show a table of your metadata file (mapping file).

*If you do not have header, you might start your first row with #

```{r warning=FALSE}
head(meta)
```

###STEP4: Construct sample_data-class using imported metadata

```{r}
sampleData <- sample_data(meta)
```

###STEP5: Import OTU table

OTU table from Mojave 16S data is “MC2017FC.otu_table.txt”.
MC-Mojave Crust-2017-FC-Fungi completed

```{r}
otus <- read.table("Mojave2020_KKelly202307.otu_table.txt",header=T,sep="\t",row.names=1)
otumat <- as(as.matrix(otus), "matrix")
OTU = otu_table(otumat, taxa_are_rows = TRUE)
```

Check imported OTU table

```{r warning=FALSE}
head(OTU)
```

###STEP6: Import taxonomy table
Taxonmy table generated from AMPtk need to be rearranged using following script.

“perl rdp_taxonmy2mat.pl<Input_taxonmy.txt>Output_taxonomy.txt”

rdp_taxonomy2mat.pl was created by Professor Jason E. Stajich

```{r}
taxmat <- read.table("Mojave2020_KKelly202307.ASVs.taxonomy.fix.txt", header=T,sep="\t",row.names=1)
taxmat <- as(as.matrix(taxmat),"matrix")
TAX = tax_table(taxmat)
```

###STEP7: Import phylogenetic tree
Phylogenetic tree can also be include for further phylogenetic analysis.

```{r warning=FALSE}
treefile = "Mojave2020_KKelly202307.ASVs.tree.phy"
tree = read.tree(treefile)
```

###STEP8: Construct Phyloseq object
To construct phyloseq object, otu table, taxonomy table, and sampleData are required. Phylogenetic tree can be included, but it is not necessary for constructing phyloseq object.
Construct Phyloseq object called "Physeq"

```{r warning=FALSE}
physeq = phyloseq(OTU,TAX,sampleData,tree)
```

Check phyloseq object
This should indicate that your physeq is a "phyloseq-class experiment-level object""

```{r warning=FALSE}
physeq
```

###STEP9: Remove singletons
Remove any OTUs that present only one time.

```{r }
physeq.prune = prune_taxa(taxa_sums(physeq) > 1, physeq)
```

```{r warning=FALSE}
physeq.prune
```

###STEP10: Plot read counts to check dataset
Check read counts: any samples that have very low reads should be removed.
[Ref](http://evomics.org/wp-content/uploads/2016/01/phyloseq-Lab-01-Answers.html)

```{r}
readcount = data.table(as(sample_data(physeq.prune), "data.frame"),
                 TotalReads = sample_sums(physeq.prune), 
                 keep.rownames = TRUE)
setnames(readcount, "rn", "SampleID")

#For plotting, use command below.
#SeqDepth = ggplot(readcount, aes(TotalReads)) + geom_histogram() + ggtitle("Sequencing Depth")
```

TotalReads of all the samples can be in this table (select only SampleID and TotalReads columns).
In order to check samples with low number of reads, "order()" can be used to sort "TotalReads" column.
In this dataset, N55.Rhizo has very low number of reads, so will will filter this sample out using the next minimum number of reads.
```{r}
readcount = readcount[order(readcount$TotalReads), c("SampleID", "TotalReads")]
```

```{r}
head(readcount)
```

```{r warning=FALSE}
set.seed(1)
physeq.prune.rarefy = rarefy_even_depth(physeq.prune, sample.size = 7909, replace = FALSE, trimOTUs = FALSE)
physeq.prune.rarefy
```

###STEP11: Glomeromycete trees

Subset taxa in phyloseq object to only include Glomeromycetes
```{r}
physeq.prune.rarefy.Glomeromycetes <- subset_taxa(physeq.prune.rarefy, Class == "Glomeromycetes")
physeq.prune.rarefy.Glomeromycetes
```

Check the taxonomy table
```{r}

tax_table(physeq.prune.rarefy.Glomeromycetes)

```
Check the phylogenetic tree for glomeromycetes
```{r}

phy_tree(physeq.prune.rarefy.Glomeromycetes)

```

Plot tree using phyloseq using genus as tip labels
```{r}
p1 <- plot_tree(physeq.prune.rarefy.Glomeromycetes, color="Substrate", shape="Site", label.tips="Genus", size="abundance", plot.margin=0.3)
p1
```
Plot a test tree using ggtree
```{r}
# Plot the modified tree
p2 <- ggtree(phy_tree(physeq.prune.rarefy.Glomeromycetes), layout="rectangular") +
  geom_tiplab(size=2, color="black")
#p2

```

Extract species names from the taxonomy table for tip names, replace unknown species with "Unknown AMF"
```{r}
# Extract species names from the taxonomy table
species_names <- tax_table(physeq.prune.rarefy.Glomeromycetes)[, "Genus"]
# replace with unknown AMF if species name is missing
species_names[is.na(species_names)] <- "Unknown AMF"
```
Match phylogeny tree tips with species names from taxonomy table
```{r}
# Get the tip labels (ASV IDs) from the phylogenetic tree
tip_labels <- phy_tree(physeq.prune.rarefy.Glomeromycetes)$tip.label

# Match ASV IDs with species names and replace tip labels
new_tip_labels <- species_names[match(tip_labels, rownames(species_names))]

new_tip_labels
```
Replace phylogenetic tree tip labels with species names combined with ASV number, with a space between
```{r}

object <- phy_tree(physeq.prune.rarefy.Glomeromycetes)

tiplabels <- object$tip.label

new_tip_labels <- paste(tiplabels, new_tip_labels, sep = " ")

object$tip.label <- new_tip_labels

object
```

Add color pallettes for heatmap
```{r}
#OKeeffe <- c("#796B58", "#1C1A18", "#D2BAA4", "#95D1E6", "#96B5BC","#BB8677","#DCD0CB", "#1D633F")
pal <- c("forestgreen", "yellow2", "red",  "darkorange")
```
Based on NCBI type species from NCBI reference phylogeny (see ./NCBI_ref_phylo), name the clades
```{r}
# Plot the modified tree with species names as tip labels
clades <- data.frame(node=c(182, 277, 158), Taxonomy=c("Rhizophagus", "Glomus", "Funelliformis"))
```
drop tips that are too long to be true AMF
```{r}
to_drop <- c("ASV4995 Gigaspora", "ASV6071 Unknown AMF", "ASV7102 Gigaspora", "ASV5616 Gigaspora")

pgroup <- drop.tip(object, to_drop)
```

Plot the tree with clade names
```{r}
p3 <- ggtree(pgroup, layout="rectangular") + geom_tiplab(size = 1, align = TRUE, linesize = .1, offset = .38, linetype = 0, color = "black")  +
  geom_cladelab(node=174, label="Rhizophagus", fontsize=2) +
  geom_cladelab(node=269, label="Glomus", fontsize=2) +
  geom_cladelab(node=168, label="Funelliformis", fontsize=2) +
  geom_cladelab(node=151, label="Glomus group 2", fontsize=2) + 
  geom_cladelab(node=171, label="Septoglomus", fontsize=2)
  p3
```
Plot tree with tip labels and save (Figure S2)
```{r}
ptest <- ggtree(pgroup, layout="rectangular",size=1,aes(color=as.numeric(label))) +
  scale_color_gradient("Bootstrap", low = "green", high = "black", limits=c(0,1)) + 
  geom_text2(aes(subset=!isTip, 
label=node)) +
  geom_tiplab(size=2, color="black") + theme(legend.position = "bottom")

ptest

```

Aave as png and SVG
```{r}
ggsave("glomeromycete_test.png", plot = ptest, width = 8, height = 5)
#182, 277, 158 are nodes to label
```

```{r}
svg("./Supplemental_phylogeny.svg", width = 8, height = 5)
ptest
dev.off()
```
Make color pallette into color gradient
```{r}

# Define the number of colors you want in the palette
num_colors <- 100

# Generate the palette from white to red
palette_white_to_red <- colorRampPalette(c("black", "yellow1", "darkorange","red"))(num_colors)

# Display the first few colors in the palette
head(palette_white_to_red)
```


```{r}
# Get the OTU table
otu_table <- otu_table(physeq.prune.rarefy.Glomeromycetes)

species_names[is.na(species_names)] <- "Unknown Glomeromycete"

# Replace row names in the OTU table with species names
rownames(otu_table) <- new_tip_labels

# Get the sample data from the phyloseq object
sample_data <- as.data.frame(sample_data(physeq.prune.rarefy.Glomeromycetes))

# Extract the variables you want to use as column names (e.g., soil type, site)
column_names <- paste(sample_data$Substrate, sample_data$Site, sep = " ")

#log_otu_table <- log10(otu_table + 1)  # Adding 1 to avoid log(0)

# Replace column names in the OTU table
colnames(otu_table) <- column_names
#log_otu_table
```
Create log normalized dataframe for heatmap 
```{r}

#colnames(log_otu_table) <- make.unique(colnames(log_otu_table))

df_otu <- data.frame(otu_table, check.names=FALSE)

df_otu <- t(rowsum(t(df_otu), group = colnames(df_otu), na.rm = T))

log_df_otu <- log10(df_otu + 1)
#df_otu
```
Plot the heatmap using ggtree
```{r}


p_heatmap <- gheatmap(p3, log_df_otu, width=1.3, offset=.13, 
                      legend_title="Relative Abundance", colnames_position="bottom", 
                      colnames_angle=45, colnames_offset_y=0, hjust=1, font.size=1.25) +
  scale_fill_gradientn("Abundance", colours = palette_white_to_red) +
  scale_y_continuous(limits=c(-10, 149)) + theme_tree(legend.position="right") #+ geom_tiplab(size = .8, align = TRUE, #linesize = .25, offset = 1, linetype = 0, color = "grey")
plot(p_heatmap)
```

Add phylopic to the plot
```{r}

#cat <- get_phylopic("23cd6aa4-9587-4a2e-8e26-de42885004c9")
img <- pick_phylopic(name = "Glomus diaphanum", n = 1)

d <- p_heatmap + add_phylopic(img, x = .05, y = 120,
                        ysize = 25, color = "black") +
  theme(plot.title = element_text(hjust = 1),
        legend.text = element_text(hjust = 0.5)) # Center legend text
d
```
Save
```{r}
svg("./phyloheatmap.svg", width = 8, height = 5)
d
dev.off()
```

