---
title: "Mojave_Colorado_Div_ITS_phyloseq"
author: "Kian Kelly"
date: "Updated on December 21, 2023"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###STEP1: Load all necessary packages

```{r warning=FALSE, message=FALSE}
library(ape)
library(vegan)
library(dplyr)
library(scales)
library(grid)
library(reshape2)
library(phyloseq)
library(magrittr)
library(ggplot2)
library(ggpubr)
library(plyr)
library(data.table)
library(tidyr)
library(tidyverse)
library(multcompView)
library(MoMAColors)
library(patchwork)

```

###STEP2: Import Mapping file (metadate file) 

Kian notes:

I excluded Nat's samples from this analysis by removing them from the metadata file. 

```{r}
meta = read.table("metadata.tsv",header=TRUE,row.names=1,sep="\t",stringsAsFactors=FALSE)
```

###STEP3: Check if your metadata file has been imported successfully 

```{r warning=FALSE}
head(meta)
```

###STEP4: Construct sample_data-class using imported metadata

Kian notes: basically makes a phyloseq version of your metadata file
```{r}
sampleData <- sample_data(meta)
```

###STEP5: Import OTU table
Kian notes: OTU table here combines my data with Nat's.
NOTE: IF THE HEADER IS IMPROPERLY READ, REMOVE THE COMMA ADDED BY PHYLOSEQ!!!
```{r}
getwd()
otus <- read.table("Mojave2020_KKelly202307.otu_table.txt",header=T,sep="\t",row.names=1)
otumat <- as(as.matrix(otus), "matrix")
OTU = otu_table(otumat, taxa_are_rows = TRUE)
```

Check imported OTU table

```{r warning=FALSE}
head(OTU)
```

###STEP6: Import taxonomy table Taxonmy 
Note: table generated from AMPtk needs to be rearranged using a perl script.

"perl rdp_taxonmy2mat.pl\<Input_taxonmy.txt\>Output_taxonomy.txt"

rdp_taxonomy2mat.pl was created by Professor Jason E. Stajich

```{r}
taxmat <- read.table("Mojave2020_KKelly202307.ASVs.taxonomy.fix.txt", header=T,sep="\t",row.names=1)
taxmat <- as(as.matrix(taxmat),"matrix")
TAX = tax_table(taxmat)
head(TAX)
```

###STEP7: Import phylogenetic tree

```{r warning=FALSE}
#treefile = "Mojave2020_KKelly202307.ASVs.tree.phy"
#tree = read.tree(treefile)
```

###STEP8: Construct Phyloseq object 

To construct phyloseq object, otu
table, taxonomy table, and sampleData are required. Phylogenetic tree
can be included, but it is not necessary for constructing phyloseq
object. Construct Phyloseq object called "Physeq"

```{r warning=FALSE}
sample_names(OTU)
sample_names(TAX)
sample_names(sampleData)


physeq = phyloseq(OTU,TAX,sampleData)
```

Check phyloseq object This should indicate that your physeq is a
"phyloseq-class experiment-level object""

```{r warning=FALSE}
physeq
```

###STEP9: Remove singletons Remove any OTUs that present only one time.

```{r }
physeq.prune = prune_taxa(taxa_sums(physeq) > 1, physeq)
```

```{r warning=FALSE}
physeq.prune
```
```{r}
readcount = readcount[order(readcount$TotalReads), c("SampleID", "TotalReads")]
```

```{r}
head(readcount)
```

The following code rarefys the data. Make sure the sample size is equal to the sample with the lowest read count. 
```{r warning=FALSE}
set.seed(1)
physeq.prune.rarefy = rarefy_even_depth(physeq.prune, sample.size = 7909, replace = FALSE, trimOTUs = FALSE)
physeq.prune.rarefy
```

```{r}
physeq.prune.rarefy.SF = subset_samples(physeq.prune.rarefy, Layer=="Surface")
```

### subset only for Fungi
```{r}

#physeq.prune.rarefy = subset_taxa(physeq.prune.rarefy, Kingdom == "Fungi")
#physeq.prune.rarefy.SF = subset_taxa(physeq.prune.rarefy, Kingdom == "Fungi")

```


# Observed alpha diversity

```{r}
physeq.prune.rarefy.SF = subset_samples(physeq.prune.rarefy, Layer=="Surface")
sample_data(physeq.prune.rarefy.SF)$Site = factor(sample_data(physeq.prune.rarefy.SF)$Site, levels = c("CIMA", "ODLO", "AB", "GMT","TP", "TP2"))
```

```{r}
alpha.diversity.SF = estimate_richness(physeq.prune.rarefy.SF, measures = c("Observed"))
data.anova.SF = cbind(sample_data(physeq.prune.rarefy.SF), alpha.diversity.SF)
physeq.prune.rarefy.anova.SF = aov(Observed ~ Site, data.anova.SF)
summary(physeq.prune.rarefy.anova.SF)
```


```{r}
tukey.ps = TukeyHSD(x=physeq.prune.rarefy.anova.SF, 'Site', conf.level = 0.95)

generate_label_df <- function(tukey.ps, variable){
  
  # Extract labels and factor levels from Tukey post-hoc 
  Tukey.levels <- tukey.ps[[variable]][,4]
  Tukey.labels <- data.frame(multcompLetters(Tukey.levels)['Letters'])
  
  #I need to put the labels in the same order as in the boxplot :
  Tukey.labels$treatment=rownames(Tukey.labels)
  Tukey.labels=Tukey.labels[order(Tukey.labels$treatment) , ]
  return(Tukey.labels)
}

LABELS=generate_label_df(tukey.ps, 'Site')
names(LABELS) = c('Letters','Site')
```

```{r}
LABELS
```

```{r}
ylabel <- data.frame("ylabel" = c(700,1250,500,700,650,1000))
LABELS$ylabel<-ylabel$ylabel
LABELS
```
```{r}
physeq.prune.rarefy.plot.richness.Site = plot_richness(physeq.prune.rarefy.SF, x="Site", color=("Site"), measures=c("Observed")) + geom_boxplot(lwd=0.5) + stat_compare_means(method = "anova", label.y = 1500) + theme(plot.title = element_text(hjust = 0.5)) + geom_text(data=LABELS, aes(x=LABELS$Site, y = LABELS$ylabel , label = LABELS$Letters)) + theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_color_manual(values = Dali) 
physeq.prune.rarefy.plot.richness.Site
```

# Black yeast
```{r}
physeq.prune.rarefy.black = subset_taxa(physeq.prune.rarefy, Order=="Chaetothyriales" | Order=="Capnodiales" | Order=="Dothideales" | Order=="Pleosporales" )
physeq.prune.rarefy.black
```

```{r}
physeq.prune.rarefy.SF.black = subset_samples(physeq.prune.rarefy.black, Layer=="Surface")
```

##Make taxonomy table into a matrix and relabel NA as unknown
```{r}
tax.fun <- as(tax_table(physeq.prune.rarefy.SF.black),"matrix")
head(tax.fun)
tax.fun[is.na(tax.fun)] <- "Unknown"
head(tax.fun)
```

###Convert tax table back to phyloseq object and generate phyloseq object with new tax table
```{r}
TAX.fun <- tax_table(tax.fun)
fun.3 <- phyloseq(sample_data(physeq.prune.rarefy.SF.black),otu_table(physeq.prune.rarefy.SF.black),TAX.fun)
fun.3
```

```{r}
glom.fun <- tax_glom(fun.3,taxrank = "Genus")
glom.fun
```

```{r}
tax_table(glom.fun)
```

##Transform OTU table to show relative abundance
##Samples can also be merged together by a variable in the mapping file

```{r}
fun.abund <- merge_samples(glom.fun, "Site")
sample_data(fun.abund)$Site <- factor(sample_names(fun.abund))
fun.abund = transform_sample_counts(fun.abund, function(x) x / sum(x))
fun.abund
```

##Merge taxonomic data with OTU table and mapping file (Can alter code to change taxonomic rank to Order, Genus, Family, etc.) and change Phylum column from a factor to a character.

```{r}
data_glom.fun <- psmelt(fun.abund)
data_glom.fun$Genus <- as.character(data_glom.fun$Genus)
```

##If a phylum has less than 1% abundance, phylum name is changed to <1% abund.

```{r}
data_glom.fun$Genus[data_glom.fun$Abundance < 0.01] <- "<1% abund."
```

Count the levels present in the Phylum column

```{r}
Count = length(unique(data_glom.fun$Genus))
Count
```

Print out unique phyla names for insertion into barplots in next step.

```{r}
unique((data_glom.fun$Genus))
```

Create levels of phyla represented in barplot. Levels appear on the barplot in the order they are listed

```{r}
data_glom.fun$Genus <- factor(data_glom.fun$Genus, levels = c("Knufia", "Alternaria", "Phoma", "Unknown", "Cladophialophora", "Cladosporium", "Herpotrichia", "Pyrenophora", "Exophiala", "Epicoccum", "Phaeosphaeria", "Tumularia","Didymella","Neodevriesia", "<1% abund."))
```

###Create barplot of phyla - use facet_grid to separate samples by a variable ie in mapping file. Choose colors for plot in scale_fill_manual. color names can be found at http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf *amount of colors must match amount of level in phyla column
```{r}
data_glom.fun$Substrate = factor(data_glom.fun$Substrate, levels = c("Moss Crust", "Surface Sterile", "Subsurface", "Dirt"))
```
```{r fig.height=5, fig.width=8, fig.align="center"}
# The palette with black:
OKeeffe <- c("#796B58", "#1C1A18", "#95D1E6","#D2BAA4", "#96B5BC", "azure2", "azure3")
```

```{r}
# Get the Dali palette
Dali <- moma.colors("Dali")

# Add one more color to the palette
Dali <- c(Dali, OKeeffe)

# Check the updated palette
Dali
```

```{r}
data_glom.fun$Site = factor(data_glom.fun$Site, levels = c("TP", "TP2", "CIMA", "GMT", "ODLO", "AB"))
```

```{r fig.height=5, fig.width=8, fig.align="center"}
# The palette with black:
taxcom_layer_pl = ggplot(data = data_glom.fun, mapping = aes_string(x = "Site" ,y = "Abundance", fill = "Genus" )) + 
  geom_bar(stat="identity", position="fill") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_fill_manual(values = Dali)
print(taxcom_layer_pl)
taxcom_layer_pl + physeq.prune.rarefy.plot.richness.Site
```

# Regular bar plot

##Make taxonomy table into a matrix and relabel NA as unknown
```{r}
physeq.prune.rarefy.f = subset_taxa(physeq.prune.rarefy, Kingdom == "Fungi")
physeq.prune.rarefy.SF.f = subset_taxa(physeq.prune.rarefy, Kingdom == "Fungi")
```

```{r}
tax.fun <- as(tax_table(physeq.prune.rarefy.SF.f),"matrix")
head(tax.fun)
tax.fun[is.na(tax.fun)] <- "Unknown"
head(tax.fun)
```

###Convert tax table back to phyloseq object and generate phyloseq object with new tax table
```{r}
TAX.fun <- tax_table(tax.fun)
fun.3 <- phyloseq(sample_data(physeq.prune.rarefy.SF.f),otu_table(physeq.prune.rarefy.SF.f),TAX.fun)
fun.3
```

```{r}
glom.fun <- tax_glom(fun.3,taxrank = "Class")
glom.fun
```

```{r}
tax_table(glom.fun)
```

##Transform OTU table to show relative abundance
##Samples can also be merged together by a variable in the mapping file

```{r}
fun.abund <- merge_samples(glom.fun, "Climate")
sample_data(fun.abund)$Climate <- factor(sample_names(fun.abund))
fun.abund = transform_sample_counts(fun.abund, function(x) x / sum(x))
fun.abund
```

##Merge taxonomic data with OTU table and mapping file (Can alter code to change taxonomic rank to Order, Class, Family, etc.) and change Phylum column from a factor to a character.

```{r}
data_glom.fun <- psmelt(fun.abund)
data_glom.fun$Class <- as.character(data_glom.fun$Class)
```

##If a phylum has less than 1% abundance, phylum name is changed to <1% abund.

```{r}
data_glom.fun$Class[data_glom.fun$Abundance < 0.01] <- "<1% abund."
```

Count the levels present in the Phylum column

```{r}
Count = length(unique(data_glom.fun$Class))
Count
```

Print out unique phyla names for insertion into barplots in next step.

```{r}
unique((data_glom.fun$Class))
```

Create levels of phyla represented in barplot. Levels appear on the barplot in the order they are listed

```{r}
data_glom.fun$Class <- factor(data_glom.fun$Class, levels = c("Agaricomycetes", "Dothideomycetes", "Pezizomycetes", "Unknown", "Microbotryomycetes", "Eurotiomycetes","Sordariomycetes", "Saccharomycetes", "Rhizophlyctidomycetes", "Tremellomycetes","Glomeromycetes","Lecanoromycetes","Leotiomycetes","Basidiobolomycetes","<1% abund."))
```

###Create barplot of phyla - use facet_grid to separate samples by a variable ie in mapping file. Choose colors for plot in scale_fill_manual. color names can be found at http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf *amount of colors must match amount of level in phyla column

```{r}
data_glom.fun$Climate = factor(data_glom.fun$Climate, levels = c("Mojave_Desert", "Colorado_Desert", "Coastal"))
```

### Remove NA again
```{r}
data_glom.fun[is.na(data_glom.fun)] <- "Unknown"
```

```{r fig.height=5, fig.width=8, fig.align="center"}

taxcom_layer = ggplot(data = data_glom.fun, mapping = aes_string(x = "Climate" ,y = "Abundance", fill = "Class" )) + 
  geom_bar(stat="identity", position="fill") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + scale_fill_manual(values = Dali)
taxcom_layer_pl + physeq.prune.rarefy.plot.richness.Site + taxcom_layer
```


###STEP11: Beta diversity
```{r}
ps.dist = phyloseq::distance(physeq.prune, "bray")
```

### Prune taxa with low counts, subset for the layer, ordinate
```{r}
physeq.prune.x = prune_taxa(taxa_sums(physeq.prune.rarefy) > 10, physeq.prune.rarefy)
physeq.prune.x.SF = subset_samples(physeq.prune.x, Layer=="Surface")
physeq.prune.x.SF.ord <- ordinate(physeq.prune.x.SF, "PCoA", "bray")
```

```{r}
sample_data(physeq.prune.x)$Substrate = factor(sample_data(physeq.prune.x)$Substrate, levels = c("Moss Crust","Subsurface", "Dirt", "Surface Sterile"))
```

###Beta by Climate

```{r}

# Perform ordination
physeq.prune.x.SF.ord.2 <- ordinate(physeq.prune.x.SF, "PCoA", "bray")

```

```{r}

ps.dist2 = phyloseq::distance(physeq.prune.x.SF, "bray")

adonis2(ps.dist2 ~ Climate, as(sample_data(physeq.prune.x.SF),"data.frame"))

```



```{r}
sample_data(physeq.prune.x)$Climate = factor(sample_data(physeq.prune.x)$Climate, levels = c("Mojave_Desert","Colorado_Desert", "Coastal"))
```

```{r}
psClimate = plot_ordination(physeq.prune.x.SF, physeq.prune.x.SF.ord, type = "Climate", color = "Climate")  + theme_bw() + theme(plot.title = element_text(hjust = 0.5)) + stat_ellipse(geom = "polygon", alpha = 1/12) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + geom_point(size=5) + scale_colour_manual(values = Dali) + annotate("text", x = .5, y = 0.7, label = "PERMANOVA, p = 0.001")

combined <- psClimate + physeq.prune.rarefy.plot.richness.Site + taxcom_layer_pl + taxcom_layer + plot_annotation(tag_levels = 'A')
combined 
```

```{r}
png("./multipannelvenn.png", width = 16, height = 12, units = "in", res = 400)
combined 
dev.off()
```









